---
title: "Wrapping my head around circular means"
author: "Ruan van Mazijk"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For the first proper blog-post on this blog, I thought I'd share a recent statistical problem I've had as part of my Masters work.

One of my chapters is about comparing patterns of growth and flowering phenology (the timing of these sorts of events during the year) across a few dozen species of sedge (Cyperaceae)---gramminoid plants common here in the fynbos in the Cape Floristic Region.

For one analysis, I am comparing flowering data between six clades. The biological details are not the focus of this post. I am more interested (for the moment) in sharing the analytical approach I've learnt.

Normally in ecology and related disciplines, when we wish to compare two or more distributions of values, we would employ a $t$-test or _ANOVA_. This is all very well if the values we wish to compare fall on a number line.

This is not the case with phenological or seasonal data.

# A small introduction to circular data

A variable describing the time of year, time of day or season of an event is an example of _circular data_. A simple example of could involve one trying to estimate when, on average, people sleep.


```{r sleep-example-1}
sleeping <- c(
  24, 23, 24, 23, 22, 24, 23, 21, 22, 24, 23, 20, 22, 23, 24, 24,
  1, 2, 2, 1, 3, 2, 2, 2, 2, 3, 4, 4, 3, 3, 4, 5, 5, 6, 3, 2, 1
)
hist(sleeping,
  breaks = 24,
  xlab = "Time of day (hour)", ylab = "No. people sleeping",
  main = ""
)
```

From this made-up data it is clear that people sleep at night. But our go-to statistic, the mean, fails at describing this.

```{r sleep-example-2}
mean(sleeping)
```

Nobody was asleep at 11h30 in the afternoon! How do can we
