---
title: "Wrapping my head around circular means"
author: "Ruan van Mazijk"
output: github_document
fig_caption: true
bibliography: "2019-03-25-circular-data.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For the first proper blog-post on this blog, I thought I'd share a recent statistical problem I've had as part of my Masters work.

One of my chapters is about comparing patterns of growth and flowering phenology (the timing of these sorts of events during the year) across a few dozen species of sedge (Cyperaceae)---gramminoid plants common here in the fynbos of the Cape Floristic Region.

For one analysis, I am comparing flowering data between six clades. The biological details are not the focus of this post. I am more interested (for the moment) in sharing the analytical approach I've learnt.

Normally in ecology and related disciplines, when we wish to compare two or more distributions of values, we would employ a $t$-test or _ANOVA_. This is all very well if the values we wish to compare fall on a number line.

This is not the case with phenological or seasonal data.

## A small introduction to circular data

A variable describing the time of year, time of day or season of an event is an example of _circular data_. A simple example of could involve one trying to estimate when, on average, people sleep:

```{r sleep-example-plot-linear, fig.cap="Histogram of sleeping times."}
# Make up some data
sleeping <- c(
  24, 23, 24, 23, 22, 24, 23, 21, 22, 24, 23, 20, 22, 23, 24, 24,
  1, 2, 2, 1, 3, 2, 2, 1, 1, 3, 4, 2, 3, 3, 4, 5, 5, 6, 3, 1, 1
)
# Plot its distribution
hist(sleeping,
  breaks = 24,
  xlab   = "Time of day (hour)", ylab = "No. people sleeping",
  main   = ""
)
```

From this made-up data it is clear that people sleep at night. But our go-to statistic, the mean, fails at describing this:

```{r sleep-example-mean}
mean(sleeping)
```

Nobody was asleep at 11h30 in the afternoon! This statistic is not very helpful. How can we better describe the average time of asleepness?

Part of this issue is that time-of-day is not a _linear_ measurement, but a _circular_ one. It does not fall on a number line, as 24h00 is the same "thing" as 00h00. Our sleeping data would be better visualised as follows (using the R-package [`ggplot2`](https://ggplot2.tidyverse.org/)) [@ggplot2]:

```{r sleep-example-plot-circular, warning=FALSE, fig.cap="Circular histogram of sleeping times."}
library(ggplot2)
# Set theme to something simpler, cleaner
theme_set(theme_minimal())
ggplot(as.data.frame(sleeping), aes(sleeping)) +
  geom_histogram(bins = 24) +  # because 24hrs in a day
  scale_x_continuous(
    breaks = seq(0, 24, 2), 
    labels = paste0(seq(0, 24, 2), "h00")  # 01h00, 02h00 etc.
  ) +
  # Make the plot circular
  coord_polar() +
  theme(
    axis.title.x = element_blank(),
    # Remove non-sensical y-axis details now that plot is circular
    axis.title.y = element_blank(),
    axis.text.y  = element_blank()
  )
```

Here, it is much clearer that people tend to be asleep around 00h00 to 01h00. The R-package [`circular`](https://cran.r-project.org/web/packages/circular/index.html) [@circular] provides an interface to this common form of data in the social sciences and ecology. It enables us to calculate statistics such as the _circular mean_, which accounts for the fact that our data live on a circle and not a number line.

To convert data into explicitly circular data, we use `circular::as.circular()`, which takes a vector as input:

```{r sleep-example-circular-pkg, fig.cap="The default plot produced by the `plot.circular()` method."}
library(circular)
sleeping <- as.circular(sleeping, 
  type     = "directions",
  units    = "hours",
  template = "clock24"
)
sleeping
plot(sleeping)
```

Now, the mean makes much more sense:

```{r sleep-example-circular-mean, fig.cap="The mean is now more sensible, thanks to `circular::`."}
mean(sleeping)
plot(sleeping)
points(mean(sleeping), col = "red")
```

## My data: flowering times in Cape sedges

I began this post by saying that I wanted to share the application of circular statistics to some data for my Masters (albeit a small example thereof).

As above, I am comparing flowering data between six clades of gramminoid fynbos plants (Cyperaceae, Tribe Schoeneae), namely the genera _Schoenus_ L. and _Tetraria_ Beauv. _Tetraria_ is known to be polyphyletic, and until recently the only _Schoenus_ species in southern Africa was the cosmopolitain _S. nigricans_ L. The rest, used below, have been transferred from _Tetraria_ Beauv. to _Schoenus_ L. by @elliott2018.

The last large-scale conspectus of the Cape flora was by @manning2012. The flowering data I use here are from that flora, with species names ammended to match then current taxonomy [as in @elliott2018]. Each species has flowering times as follows:

```{r setup-flowering-data, include=FALSE}
library(here)
library(tidyverse)
flowering_times <- 
  read_csv(here(
    "assets/data",
    "CFR-Schoeneae-flowering-times_from-Manning-and-Goldblatt-2012_rvm.csv"
  )) %>%
  mutate(species = ifelse(is.na(new_name_rvm), species, new_name_rvm)) %>%
  rename(clade = clade_rvm) %>%
  select(clade, species, flowering_start, flowering_end) %>%
  arrange(clade) %>%
  filter(!is.na(clade))
```

```{r print-flowering-data, echo=FALSE}
flowering_times %>%
  select(-clade) %>%
  knitr::kable(
    col.names = c("Species", "Start", "End"),
    caption = paste(
      "Flowering times (start and end) for _Schoenus_ L. and _Tetraria_ Beauv.",
      "(Cyperaceae, Tribe Schoeneae), from @manning2012."
    )
  )
```

Along with these data, taxonomic research groups these species into three clades in each genus (_Schoenus_ and _Tetraria_). That merely explain the colour-coding used in the figures below.

For now, let's unpack how to take `"Jan"`, `"Feb"`, `"Mar"` and turn it into meaningful numerical data, and treat that data circularly such that December and January are adjecent.

### Flowering months as numerical data

First, we need to convert the monthly abbreviations into numbers. Simple enough: `"Jan"` = 1, `"Feb"` = 2, etc. I apply this process for each rows' `flowering_start` and `flowering_end` values, row-by-row, using [`purrr`]()--a functional programming package for R.

```{r get-month-int}
get_month_int <- function(month_abb) {
  # Converts "Jan", "Feb", etc. -> 1, 2, etc.
  month_abb %>%
    map_int(~ ifelse(.x %in% month.abb,
      which(month.abb == .x),
      NA
    ))
}
# Apply get_month_int() to the data
flowering_times <- flowering_times %>%
  mutate(
    flowering_start = get_month_int(flowering_start),
    flowering_end   = get_month_int(flowering_end)
  )
flowering_times
```

Now, let's create sequences of months that a species is in flower for, starting at `flowering_start`, ending at `flowering_end`. This will be embedded within the dataframe as a [_list-column_]()--where each entry in the column is itself a list, nested in the dataframe. I apply this process for each pair of `flowering_start`-`flowering_end` values, again, using [`purrr`]()--a functional programming package for R.

```{r get-month-seq}
get_month_seq <- function(start, end) {
  # Creates seq from start to end, unless end < start
  map2(start, end, function(.start, .end) {
    if (any(is.na(c(.start, .end)))) {
      return(NA)
    } else if (.start > .end) {
      return(c(.start:12, 1:.end))  # e.g. Dec->Apr = 12, 1, 2, 3, 4
    } else if (.start <= .end) {
      return(.start:.end)
    }
  })
}
flowering_times <- flowering_times %>%
  mutate(flowering_months = get_month_seq(flowering_start, flowering_end))
flowering_times
```

Lastly, we just need to reshape this data into a form `ggplot2` can understand: long-form data.

```{r reshape-flowering-data}
flowering_times <- flowering_times %>%
  unnest() %>% # unnest the list-columns so ggplot can understand
  # Arrange species by sub-generic clade
  arrange(clade) %>%
  mutate(species = factor(species, levels = unique(species)))
flowering_times
```

### Plotting flowering times

Now we can really plot this data:

```{r plot-flowering-times-linear}
flowering_plot <- 
  ggplot(flowering_times, aes(flowering_months, species, fill = clade)) +
    geom_tile() +
    scale_x_continuous(breaks = 1:12, labels = month.abb) +
    theme(
      axis.title.x = element_blank(),  # no need for x-axis title
      axis.text.y  = element_text(hjust = 0)  # left-justify species names
    )
flowering_plot
```

It's clear that all the (austral) summer flowering species (`"Dec"`--`"Feb"`) could really benefit from a circular plot. Thus:

```{r plot-flowering-times-circular}
flowering_plot +
  # Make the plot circular
  coord_polar() +
  # Remove non-sensical y-axis details now that plot is circular
  theme(
    axis.text.y  = element_blank(),
    axis.title.y = element_blank()
  )
```

This is much more realistic. Problem is, it squashes up all the species in the centre of the circle.

Perhaps a better summary would be a count of the number of species in flower in a given month, for each clade. Like so:

```{r plot-flowering-times-circular-bars}
flowering_times %>%
  filter(clade != "S. nigricans") %>%  # only 1 species, remove for this plot
  ggplot(aes(flowering_months, fill = clade)) +
    geom_bar() +
    scale_x_continuous(breaks = 1:12, labels = month.abb) +
    coord_polar() +
    facet_wrap(~clade) +
    theme(
      axis.title.x    = element_blank(),
      axis.text.y     = element_blank(),
      axis.title.y    = element_blank(),
      strip.text      = element_blank()  # legend duplicates facet labels
    )
```

Now that we've figured out how to plot the data in some useful ways, let's now try and analyses with the appropriate statistics.

### Statistics for flowering times

First, let's just simply our dataset for the purposes of this exercise. I'm going to just group months where _any_ species is in flower, grouped by clade:

```{r summarise-flowering-data}
flowering_summary <- flowering_times %>%
  select(clade, flowering_months) %>%
  # Split the dataframe into a list according to clade
  split(.$clade) %>%
  # For each item in the list, return only the vector of flowering months
  map(pull, flowering_months) %>%
  # Tidy up the list's names
  set_names(names(.) %>%
    # Replace hyphens ("-") & spaces (" ") with underscores ("_")
    str_replace_all("[-\ ]", "_") %>%
    # Remove fullstops (".")
    str_remove_all("\\.")
  )
flowering_summary
```

Now let's `circular::`-ise the data:

<!--FIXME: from here down-->

```{r circularise-flowering-data}
flowering_circular <- flowering_summary %>%
  map(function(.x) .x * 2) %>% 
  map(as.circular,
    type = "directions",
    units = "hours", template = "clock24",
    rotation = "clock"
  )
flowering_circular
```

Let's plot that with `plot.circular()` just to check that it worked:

```{r plot-circularised-flowering-data}
# For each item in the list (.x) and its name (.y):
imap(flowering_circular, ~plot(.x, main = .y))
```

Note, where this is more than one data point (i.e. species), the points on this plots overlap.

Let's try plotting the means now:

```{r plot-circularised-flowering-data-w-means}
# What are the means?
map(flowering_circular, mean, na.rm = TRUE)
imap(flowering_circular, function(.x, .y) {
  plot(.x, main = .y)
  points(mean(.x, na.rm = TRUE), col = "red")
})
```

## Session info

I present here my machine's state at time of writing, for reproducibility's sake.

```{r sessionInfo}
sessionInfo()
```

## Further reading

## References
